%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FILE   : main.tex
% AUTHOR : (C) Copyright 2024 by Peter Chapin
% SUBJECT: Document describing the solar system simulator.
%
% Send comments or bug reports to:
%
%       Peter Chapin
%       Computer Information Systems Department
%       Vermont State University
%       Williston, VT 05061
%       spicacality@kelseymountain.org
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%+++++++++++++++++++++++++++++++++
% Preamble and global declarations
%+++++++++++++++++++++++++++++++++
\documentclass{article}

\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{hyperref}

% The following are settings for the listings package.
\lstset{language=C,
        basicstyle=\small,
        stringstyle=\ttfamily,
        commentstyle=\ttfamily,
        xleftmargin=0.25in,
        showstringspaces=false}

\setlength{\parindent}{0em}
\setlength{\parskip}{1.75ex plus0.5ex minus0.5ex}

\newcommand{\filename}[1]{\texttt{#1}}

%++++++++++++++++++++
% The document itself
%++++++++++++++++++++
\begin{document}

%-----------------------
% Title page information
%-----------------------
\title{Solarium: A Solar System Simulator}
\author{Peter Chapin}
\date{January 7, 2024}
\maketitle

\section{Introduction}
\label{sec:introduction}

The dynamics of solar systems is an interesting research topic in astronomy. Although many solar
systems exhibit a high degree of stability even over billions of years, that is not universally
the case. When many bodies interact gravitationally, chaotic behavior can arise causing planets
to be ejected from the system or even to collide. There is evidence that this sort of chaos even
occurred in our own solar system early in its history\footnote{It is currently believed that a
Mars sized planet collided with the Earth causing the formation of the Moon.}

The purpose of this program is to simulate the motion of objects in a solar system. The
simulation is eventually intended to run for millions (or even billions) of simulated years to
explore long term solar system stability. We will assume that a solar system is composed of a
finite number of gravitationally interacting ``point'' masses. We will also consider each object
to be a sphere with a fixed size so that collisions can be detected and counted.

This model of the solar system is reasonably accurate for our purposes. The effects of
interplanetary gas and dust as well as exotic ``dark matter'' or ``dark energy'' should not be
significant on the scales of interest here. In this model the motion of the objects in the solar
system can be completely described by the response to their mutual gravitational attraction.

It should be said, that although it is hoped that the program described here will produce
reasonable and interesting results, the primary goal of this exercise is to demonstrate various
parallel programming techniques in the context of a typical high performance computing
application.

The rest of this document is organized as follows. Section~\ref{sec:computation} describes the
required computation in more detail along with some mathematical background.
Section~\ref{sec:preliminaries} discusses some preliminary matters such as what system of units
should be used and how the input and output of the computation should be prepared and handled. A
sequential program that solves the problem is presented and discussed in detail in
Section~\ref{sec:sequential}. Various issues surrounding how to parallelize the computation,
along with related programming exercises are then presented in Section~\ref{sec:parallelizing}.
Finally, the document concludes with some benchmarks, a summary of results, and so forth.

\input{Computation}
\input{Preliminaries}
\input{Sequential}
\input{Parallelizing}

\textit{TODO: Add final sections!}

\end{document}
